File-based localization of numerical perturbations in data analysis pipelines

### 2020-04-06 paper.tex
Cleaned up figures
Fixed fig 2 and 3 (now fig 1)

Data analysis pipelines are known to be impacted by operating system
updates and other infrastructural factors, presumably due to the creation,
propagation and amplification of numerical errors. However, the precise
causes of such instabilities and the path along which they propagate in the
pipelines are unclear, while this process could play a major role in the
current reproducibility crisis. We present a technique to identify the
computational processes in a pipeline that create numerical differences
when executed in different computational conditions. Our technique
leverages system-call interception to reconstruct processing and file
graphs without pipeline instrumentation. By applying our tool to the
structural pre-processing pipelines released by the Human Connectome
Project, we find that linear and non-linear registration are the cause of
most numerical instabilities in such pipelines, which confirms previous
findings.

@@

@@

\toolname identifies the processes in a pipeline that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each subject, where nodes
represent application processes and files, and edges represent read and
write file accesses (Step 1 in Figure~\ref{fig:overview-tool}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified (Step 2) and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Step 3). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Step
4). The resulting labeled graph identifies the
non-reproducible processes in the pipeline.

@@

\begin{figure*}
  \centering
    \includegraphics[width=.8\textwidth]{images/spot-diagram}
    \caption{\toolname overview.}
    \label{fig:overview-tool}
  \end{figure*}

@@

\begin{listing}
  \inputminted{bash}{"bin/example.sh"}
  \caption{Example pipeline}
  \label{listing:sample-script}
\end{listing}

\begin{figure}
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph.pdf}
  \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
  \label{fig:provenance-graph}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph-dag.pdf}
  \caption{Provenance DAG, with disambiguated transient files.}
  \label{fig:provenance-dag}
\end{subfigure}
  % \includegraphics[width=.8\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  created from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes are represented with circles, files with rectangles, and read/write accesses 
  with plain edges. 
  For visualization purposes, the process tree is represented too, with
  dashed edges. Every node in the graph is labeled using (1) a process id
  created by our reconstruction, (2) the name of the executable run by the
  process or the file name. Process 0 is the initial call to the
  \texttt{sh} interpreter; the other processes are the calls to FSL bet,
  stats and maths made in Listing~\ref{listing:sample-script}. Process 73
  is forked by process 2: it was captured by \reprozip while it did not
  appear in Listing~\ref{listing:sample-script}.
}
  \label{fig:simple_script}
\end{figure}

@@

After capturing transient files, we re-run the pipeline
step by step in Condition 2 to label processes. The output files
produced by a process in Condition 2 are compared to the ones produced
in Condition 1. If no differences are found, the process is marked as
reproducible. Otherwise, the process is marked as non-reproducible, and the
output files produced in Condition 1 are copied to Condition 2, to ensure
that differences do not propagate further in the pipeline. Processes are
instrumented transparently, through a modification of the \texttt{PATH}
variable similar to the one described previously. By default, differences
in output files are identified by comparing file checksums. Other
comparison functions can also be defined for specific file types, for
instance to ignore file headers or file sections containing timestamps.
\toolname finally creates a labeled
provenance graph highlighting non-reproducible processes.

Figure~\ref{fig:iterations} illustrates a hypothetical incremental labeling 
process for the example in Figure~\ref{fig:simple_script}. 
At step 1, process \texttt{73\#bet2} is labeled as non-reproducible (red) 
as it produces files with differences, and the files produced by \texttt{73\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{73\#bet2} in 
Condition 1. At step 2, process \texttt{75\#fslmaths} is executed and then labeled 
as reproducible (green) as it produced files without differences.
At step 3, \texttt{76\#fslstats} is also labeled as 
reproducible as it didn't produce any file.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{images/p-graph-dag-labelled.pdf}
  \caption{Illustration of the labeling approach. Non-reproducible processes are shown in red.}
  \label{fig:iterations}
\end{figure}

>>>>

% Data analysis pipelines are known to be impacted by operating system
% updates and other infrastructural factors, presumably due to the creation,
% propagation and amplification of numerical errors. However, the precise
% causes of such instabilities and the path along which they propagate in the
% pipelines are unclear, while this process could play a major role in the
% current reproducibility crisis. We present a technique to identify the
% computational processes in a pipeline that create numerical differences
% when executed in different computational conditions. Our technique
% leverages system-call interception to reconstruct processing and file
% graphs without pipeline instrumentation. By applying our tool to the
% structural pre-processing pipelines released by the Human Connectome
% Project, we find that linear and non-linear registration are the cause of
% most numerical instabilities in such pipelines, which confirms previous
% findings.

@@

\newpage

@@

\toolname identifies the processes in a pipeline that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each subject, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline.

\begin{figure*}
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.5\linewidth]{images/p-graph.pdf}
    \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
    \label{fig:provenance-graph}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.4\linewidth]{images/p-graph-dag.pdf}
    \caption{Provenance DAG, with disambiguated transient files.}
    \label{fig:provenance-dag}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
      \centering
      \includegraphics[width=.4\linewidth]{images/p-graph-dag-labelled.pdf}
     \caption{Labeled DAG comparing 2 execution conditions, showing 1 non-reproducible process.}
     \label{fig:labeled-dag} 
  \end{subfigure}
    % \includegraphics[width=.8\columnwidth]{images/provenance-graphs}
    \caption{Provenance graphs
    created from the example pipeline in
    Listing~\ref{listing:sample-script}.
    Processes are represented with circles, files with rectangles, and read/write accesses 
    with plain edges. 
    For visualization purposes, the process tree is represented too, with
    dashed edges. Every node in the graph is labeled using (1) a process id
    created by our reconstruction, (2) the name of the executable run by the
    process or the file name. Process 1 is the initial call to the
    \texttt{sh} interpreter; the other processes are the calls to FSL bet,
    stats and maths made in Listing~\ref{listing:sample-script}. Process 73
    is forked by process 2: it was captured by \reprozip while it did not
    appear in Listing~\ref{listing:sample-script}.
  }
    \label{fig:spot-example}
  \end{figure*}

@@

\begin{listing}
  \inputminted{bash}{"bin/example.sh"}
  \caption{Example pipeline}
  \label{listing:sample-script}
\end{listing}
\tristan{Ali, I added a file in Listing 1, could you re-generate the trace?}

@@

@@

After capturing transient files in the first condition, we re-run the pipeline
step by step in the second one to label processes. The output files
produced by a process in both conditions are compared.
 If no differences are found, the process is marked as
reproducible. Otherwise, the process is marked as non-reproducible, and the
output files produced in the first condition to the second one, to ensure
that differences do not propagate further in the pipeline. Processes are
instrumented transparently, through a modification of the \texttt{PATH}
variable similar to the one described previously. By default, differences
in output files are identified by comparing file checksums. Other
comparison functions can also be defined for specific file types, for
instance to ignore file headers or file sections containing timestamps.
\toolname finally creates a labeled
provenance graph highlighting non-reproducible processes.

Figure~\ref{fig:labeled-dag} illustrates a hypothetical incremental
labeling of the example in Listing~\ref{listing:sample-script}. Process
\texttt{73\#bet2} is labeled as non-reproducible (red) as it produces files
with differences. To prevent the propagation of these differences, the
files produced by \texttt{73\#bet2} in Condition 2 are replaced with the
files produced by \texttt{73\#bet2} in Condition 1. Processes
\texttt{75\#fslmaths} and \texttt{76\#fslstats} are then executed and
labeled as reproducible (green) as they produce files without differences.
